### 2020-02-17 ordinal/ordinal.Rmd
ordinal tutorial corrected

This document contains a short tutorial on fitting hierarchical
ordinal polytomous regression models using the bhsdtr package, but
there is more; It also contains some explanations, illustrated with
annotated code, of several issues related to ordinal models in
general, and Signal Detection Theory and Item Response Theory in
particular. If you are interested in SDT, IRT, or ordinal models in
general, perhaps you will find something useful, new, or even mildly
provocative here.

@@

To fit a model we have to aggregate the data (this makes the process
of sampling much faster), create the fixed- and random-effects lists,
and create the data structures and the model code required by
stan. Note that here we introduce a new parameter - eta - which
represents the average internal value. If we are going to model all
the thresholds (gamma parameter vector) then this parameter is
redundant, because a constant shift in the thresholds is equivalent to
the opposite effect in eta. The bhsdtr package will automatically fix
the first element (here the only element) of the eta fixed effects
vector at 0.

@@

Note that the two assumptions, i.e., that the thresholds are constant
across participants and latent means or samples are participant-, not
item-specific seem to be accepted in the great majority of
applications of Item Response Theory. This is reflected in the IRT
terminology: thresholds are often called "item-parameters" and latent
means or samples are often called "person-parameters". Some even say
that IRT can provide separate estimates of the true item- and
person-parameters.

@@

The neff and Rhat columns look good. Here is how the model fits (this
is a maximum likelihood fit):

@@

Now we will introduce the model which accounts for the
inter-individual differences in the pattern of the thresholds by
modelling the gamma random effects. We cannot simultaneously model the
eta random effects in this model, since all the information that can
be captured by the eta parameter (i.e., the constant shift in the
thresholds) is already captured by the fixed and the random gamma effects.

@@

```{r echo=T,results="show"}              
print(fit0, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
## neff and Rhat look good
```

The neff and Rhat columns look good. Below we translate between the
gamma and the criteria vectors (fixed effects = sample average):

@@

Are the sample average thesholds different in the two models? Off
course they are, since the models are non-linear and averaging over
anything that affects any parameter of such models is a bad idea:

@@

Now we will fit a model which is similar to an SDT model, but there
will be two important differences. We will not assume that the answer
"signal" cannot be less likely for the stimulus "signal" than for the
stimulus "noise", i.e., we will not assume that d' must be
non-negative, and we will *not* assume that the pattern of the
thresholds is the same for "noise" and "signal" stimuli. If you know
your Signal Detection Theory well, this may seem to you like a
perverse idea. If that is the case, please bear with me.

In this model each participant may have two different sets of
thresholds, one for "noise" and one for "signal" stimuli. Since the
thresholds are represented as random effects, we will account for the
fact that "random" thresholds may be correlated, and they are - if
some threshold is higher (lower) than the sample average, then the
threshold above (below) it may also be higher (lower).

@@

```{r echo=T,results="show"}                
fixed1 = list(eta = ~ 1, gamma = ~ -1 + stim)
random1 = list(list(group = ~ id, gamma = ~ -1 + stim))
## We have to make sure that stim is a factor for the separate
## intercepts and slopes parametrization to work as intended.
d2$stim = as.factor(d2$stim)
## Note that because for an ordinal model the stimulus variable (here
## the 'stim' variable) is not special, we have to include it in the
## additional variables vector
adata1 = aggregate_responses(d2, response = 'r', variables = c('id', 'stim'))
sdata1 = make_stan_data(adata1, fixed1, random1, model = 'ordinal', gamma_link = link)
model1 = make_stan_model(random1, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit1 = stan(model_code = model1, data = sdata1,
                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit1, '~/windows/temp/ordinal_fit_1.rds')
}else{
    fit1 = readRDS('~/windows/temp/ordinal_fit_1.rds')
}
## We will not look at the eta parameter since it was fixed at 0
print(fit1, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
## neff and Rhat look good
```

@@

Note the difference between the main criterion (row 4). This
difference approximates d'. Even though, as we will later see, this
model does not correctly account for the effect of the stimulus, we can
still recover this effect pretty well, because d' is equivalent to a
constant shift of all the thresholds in one predefined
direction. Shifting all the criteria to the right is equivalent to
shifting the underlying distribution of "latent values" (or evidence)
to the left, so we have to subtract the average position of the
criteria for the "signal" stimuli from the average position of the
criteria for the "noise" stimuli. Here is our first attempt:

@@

In an SDT model we assume that the stimulus class affects *only* the
d' parameter, and that both criteria (gamma) and *d'* may differ
between the participants (or items, etc.). In bhsdtr d' is modelled as
log(d') = delta, which forces it to be non-negative. This is an
important assumption of Signal Detection Theory - if this assumption
is violated, for example, because of the response reversal in some
participants or conditions, then an SDT model is not valid and a
non-trivial generalization of SDT should be used instead (see the
bhsdtr preprint for a more detailed explanation and some
references). Otherwise, this is the same general hierarchical ordinal
regression model.

@@

```{r echo=T,results="show"}
adata2 = aggregate_responses(d2, stimulus = 'stim', response = 'r', variables = c('id'))
sdata2 = make_stan_data(adata2, fixed2, random2, model = 'sdt', gamma_link = link)
model2 = make_stan_model(random2, model = 'sdt', gamma_link = link)
if(fresh_start){
    fit2 = stan(model_code = model2, data = sdata2,
                pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1', 'gamma_sd_1',
                         'delta_random_1', 'gamma_random_1',
                         ## We will need this for the plot_sdt_fit function
                         'counts_new'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit2, '~/windows/temp/ordinal_fit_2.rds')
}else{
    fit2 = readRDS('~/windows/temp/ordinal_fit_2.rds')
}
print(fit2, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1', 'gamma_sd_1'))
## neff and Rhat look good
```

So how well do the two d' estimates match? We can translate between
the delta and the d' parameters by using exponentiation. We have to do
this first, before doing anything else with the posterior samples,
because the exponential function is non-linear.

@@

The point and interval estimates match quite well, with some
notable exceptions. According to the ordinal model some credible
intervals around individual d' point estimates include 0 and one point
estimate is even below 0, which contradicts an important assumption of
the SDT model. We will address this issue later.

If the participant-specific d' estimates match, the sample average d'
should also match, right? Maybe the SDT model does not fit the data
well?

@@

```{r echo=T,results="show"}
ggplot(data.frame(delta = log(dprim2i$mean)), aes(sample = delta)) +
    stat_qq() +
    ylab('Sample quantile of log(d\') = delta random effects') +
    xlab('Theoretical normal quantile') +
    labs(title = 'SDT model')
```

The individual *delta = log(d')* estimates in the SDT model seem to be
approximately normally distributed. We cannot plot log(d') estimates
based on the ordinal model, because according to this model one
participant has a negative *d'* value. However, we can see if the *d'*
point estimates based on the ordinal model are approximately normally
distributed:

@@

They seem to be approximately normally distributed as well. The
average / fixed effect *d'* in each model is estimated as the mean of
different distributions: in the ordinal model participant-specific
shifts are assumed to be normally distributed and in the SDT model
delta = log(d') random effects are assumed to be normally
distributed. We can easily account for this by making use of the
relationship between the parameters of the normal distribution and the
associated log-normal distribution:

@@

Before we go any further, just for fun, let's see what happens if we
introduce yet another assumption of selective influence, one that is
often accepted in Item Response Theory. We will fit a hierarchical SDT
model in which we will call the thresholds "item-parameters" and we
will call *d'* a "person-parameter".

```{r echo=T,results="show"}
fixed3 = list(delta = ~ 1, gamma = ~ 1)
random3 = list(list(group = ~ id, delta = ~ 1))
adata3 = aggregate_responses(d2, stimulus = 'stim', response = 'r', variables = c('id'))
sdata3 = make_stan_data(adata3, fixed3, random3, model = 'sdt', gamma_link = link)
model3 = make_stan_model(random3, model = 'sdt', gamma_link = link)
if(fresh_start){
    fit3 = stan(model_code = model3, data = sdata3,
                pars = c('delta_fixed', 'delta_sd_1', 'gamma_fixed', 'delta_random_1',
                         ## We will need this for the plot_sdt_fit function
                         'counts_new'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit3, '~/windows/temp/ordinal_fit_3.rds')
}else{
    fit3 = readRDS('~/windows/temp/ordinal_fit_3.rds')
}
print(fit3, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1'))
## neff and Rhat look good
```

@@

At the aggregated data level the fit is not perfect, but it is not
that terrible either. What about the fit at the individual level?

@@

... we see that the simplified model not only gives biased
"person-parameter" estimates but the estimates seem more reliable than
they truly are (here by about 20%). It does not make sense to
look at the estimates of the "item-parameters" since we already know
that they have little in common with the true thresholds.

Now back to Signal Detection Theory and the assumption of selective
influence. We can at least try to test this assumption. We will not
alter the ordinal model code to account for the non-negativity of d'
(and the log-normal distribution of random d' effects) just yet. As we
already know, even without this correction the estimates of
participants' shifts are similar to the *d'* estimates.

@@

Ok, we did simplify things significantly, but it still makes sense to
take a look at the other gamma effects, because these other effects
correspond to the variability in the *pattern* of the thresholds, and
the estimates of the thresholds in our ordinal model seem perfectly
fine.

@@

Note the clear pattern for the gamma_sd_1[.,2] parameters. Even though
the *sample-average* pattern of the thresholds seems to be different
for the two stimulus classes (gamma_fixed[.,2] except for
gamma_fixed[4,2] ~ d'), this difference in pattern seems to be fairly
constant across the participants: all the gamma_sd_1[.,2] parameters
seem to be near-zero (the 95% credible intervals contain zero), except
for the gamma_sd_1[4,2] parameter, which corresponds to the
variability in *d'* values. This is *really* weird, since the pattern
of the thresholds for the "noise" stimuli clearly varies with
participants - none of the 95% HPD intervals for the gamma_sd_1[.,1]
parameters include zero: in terms of standard deviations of posterior
distributions of standard deviations (phew) most are in fact quite far
from zero.

If only we could do a better job and use a model which respects the
non-negativity assumption. Is it really that complicated? Let's index
the rows of our dataset by *i*. Because of the way the model was
parametrized...

@@

However, when stim = "signal" = 1, gamma[4]_i = gamma_fixed[4,1] +
gamma_random_1[g][4,1] + gamma_fixed[4,2] + gamma_random_1[g][4,2]. We
want this other part (the last sum which represents the difference in
gamma[4] between the "noise" and "signal" stimuli) to represent log of
average distance between the thresholds = shift = *d'*, so we have to
subtract it from gamma[4]_i, exponentiate this sum, and add it again
to gamma[4]_i.

@@

Interaction is tricky, you see. For example, if you have an additive
effect of two factors on the original scale of the dependent
variable...

@@

This means that whenever you see an interactive effect it may be an
artefact of the scale of the dependent variable! Every time I say it
it makes me feel uneasy. So, instead of examining the effects in the
gamma parameters. let's have a look at the average *criteria /
thresholds*. Notice that we have to add the effect on gamma to gamma
for stimulus = "noise" to obtain gamma for stimulus = "signal".

@@

Now we can directly test if the *pattern* of the criteria may be
different by comparing *centered* criteria vectors:
```{r echo=T,results="show"}
round(HPDinterval(as.mcmc((crit2 - apply(crit2, 1, mean)) - (crit1 - apply(crit1, 1, mean)))), 2)
```

When we center the criteria for the two classes of stimuli we no
longer see any evidence of the differences in their relative
positions. The only thing that seems to differ, and not by much, is
the main criterion, but this is confounded with *d'* in our model (I
think). Selectivity seems to hold, kind of.

This is how you *begin* to test a hierarchical SDT model. An SDT model
has other problematic assumptions. One is that the time-accuracy
tradeoff can be ignored. If participants or conditions differ in the
response time, then the differences in *d'* may not mean that
sensitivity is different, because participants may simply take more
time to respond in certain experimental conditions, which inflates the
*d'* estimate. You need something like the diffusion model to deal
with this problem. If you combine the diffusion model with ratings,
you also need something like a hierarchical diffusion model with
ratings and an order-preserving link function. Sounds like something
that could perhaps be implemented in the bhsdtr package in the future.

The other problems that need to be addressed have to do with
non-linearity and aggregation. If you fit an SDT model, or any other
non-linear model, to data aggregated over grouping factors, such as
participants or items, because of non-linearity, which does not play
well with aggregation, some or all of the point and interval estimates
will be biased, perhaps severely so, and the inference will be
invalid. You can see how horribly biased the point *and* interval
estimates of every SDT parameter may become when you overly aggregate
your data in the bhsdtr preprint. I am now in the process of examining
the direction and magnitude of bias in SDT models fitted to
overly-aggregated data by using the excellent Confidence database
created by Doby Rahnev.

@@

Unfortunately, without additional assumptions, it is impossible to
model both kinds of variability at the same time even in the
hypothetical repeated-measurement-with-memory-wipeout-scenario,
because every shift in the internal value is indistinguishable from a
uniform effect in the thresholds, i.e., a constant shift of all the
thresholds. As I hope you already realize, the required additional
assumptions that make such models identifiable under certain
conditions are necessarily *causal*.

In many situations, this is both a statistical and a conceptual
problem. In particular, when the design is purely observational -
e.g., questionnaire measurement - without additional causal
assumptions, the way of describing the process of responding to a
Likert-type item in terms of internal values and thresholds is simply
redundant. As far as the design of the study is concerned, all we can
say about the internal value is that it is relative to the thresholds
- and the position of each threshold is relative to the positions of
the remaining thresholds and to the position of the internal
value. Without the selectivity assumption, to say that John thinks
that he was not crying at all during the last two weeks and to say
that *John's* threshold for the response "a few times" is above the
amount of crying that John thinks he experienced, or decided to
reveal, is to say the same thing.

It follows that *in such situations, we do not need the internal value
parameter at all*, unless there is some factor about which we can
perhaps assume that it can *only* produce a uniform effect in the
thresholds, such as the stimulus class ("noise" or "signal") in a
binary classification task, which can *perhaps* be assumed to induce a
change (*d'*) in the internal value only; In such cases the internal
value parameter may be justified and useful because it does constrain
the model. We have obtained evidence that this assumption perhaps
holds for the analyzed dataset.

@@

It is usually not an *achievement* of Item Response Theory, it is just
an unrealistic *assumption*. The illusory nature of this achievement
becomes obvious as soon as we realize that without additional strong
causal assumptions it is impossible to distinguish between genuine
internal value effects and (dis)simulation (i.e., faking), which
conceptually corresponds to the change in the *participant*-specific,
not item-specific, decision thresholds, also known as
"item-parameters". There is nothing in the *design* of the process of
measurement by a questionnaire that makes it possible to distinguish
genuine differences in internal values from the differences in the way
of responding.

Because of random assignment of the stimulus class above chance
performance in a binary classification task cannot be faked, and so
the *d'* parameter definitely captures at least part of the effect of
the stimulus on the internal value. On the other hand, every possible
pattern of responses to a questionnaire can be faked. The only way to
distinguish the "item" and "person" effects is to *assume selective
influence*, i.e., that there is some factor f (such as the stimulus
class in a binary classification task, or, in case of a depression
questionnaire, participants trait, or a therapeutic intervention)
which can *only* affect the internal value, thus producing a
*constant* shift in the thresholds. The assumption of selective
influence makes the model with the two kinds of parameters
identifiable if there are repeated measurements for at least one
participant and one item or many participants provide single responses
to all of the items.

@@

If we campared the estimates of the main decision criterion (i.e.,
bias) in SDT models fitted to two different large samples of
participants the difference would likely be negligible with very
narrow confidence / credible intervals. By the same token, instability
of item parameters in IRT models cannot be estimated by assuming
selective influence and comparing estimates of "item-parameters"
obtained in two different large samples, because large-sample average
estimates of item parameters are the best way to *hide or misplace*
(it becomes part of the variability in the "person-parameters") the
intra- or inter-individual variability in item parameters; By assuming
that internal values are participant-specific and item parameters are
item-specific we force the estimates of item parameters to become
estimates of average person-specific thresholds. This not only begs
the question, but also, because the model is non-linear, makes the
estimates of item-parameters *asymptotically biased*.

It does not take much thought to see that the assumption of constant
"item-parameters" at the group / sample / population level is utterly
unrealistic. It follows that the achievement of separating participant
and item effects in IRT models is often not an achievement at all, it
is just wishful thinking.

The justification of the selective influence assumption comes from a
psychological theory of the task, not from the design of the
study. All that is guaranteed by randomly assigning the stimulus class
(i.e., by choosing the stimulus class at random on every trial) is
that any statistical effect of the stimulus class is an unbiased
estimate of the total causal effect of the stimulus class. To
empirically justify the assumption of *selective* influence of the
stimulus class on the internal evidence value we have to *test* if the
stimulus class does not also affect the thresholds.

>>>>

This document contains a short tutorial on fitting hierarchical
ordinal polytomous regression models using the bhsdtr package, but
there is more; It also contains some explanations, illustrated with
annotated code, of several issues related to ordinal models in
general, and Signal Detection Theory and Item Response Theory in
particular. If you are interested in SDT, IRT, or ordinal models in
general, perhaps you will find something useful here.

@@

To fit a model we have to aggregate the data as much as possible
without distorting the hierarchical structure (this makes the process
of sampling much faster), create the fixed- and random-effects lists,
and create the data structures and the model code required by
stan. Note that we introduce a new parameter - eta - which represents
the average internal value. If we are going to model all the
thresholds (gamma parameter vector) then this parameter is redundant,
because a constant shift in the thresholds is equivalent to the
opposite effect in eta. The bhsdtr package will automatically fix the
first element (here the only element) of the eta fixed effects vector
at 0, which will make this model identifiable.

@@

Note that the two assumptions, i.e., that the thresholds are constant
across participants and latent means or samples are participant-, not
item-specific seem to be often accepted in applications of Item
Response Theory. This is also reflected in the IRT terminology:
thresholds are often called "item-parameters" and latent means or
samples are often called "person-parameters".

@@

Here is how the model fits (this is a maximum likelihood fit):

@@

Now we will introduce the model which accounts for the
inter-individual differences in the pattern of the thresholds by
modelling the gamma random effects.

@@

```{r echo=T,results="show"}              
print(fit0, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
```

Below we translate between the gamma and the criteria vectors (fixed
effects):

@@

Are the average thesholds different in the two models? Off course they
are, since the models are non-linear and averaging over anything that
affects any parameter of such models is a bad idea:

@@

Now we will fit a model which is similar to an SDT model, but there
will be two important differences. We will not assume that the answer
"signal" cannot be less likely for the stimulus "signal" than for the
stimulus "noise", i.e., we will not assume that *d'* must be
non-negative, and we will *not* assume that the pattern of the
thresholds is the same for "noise" and "signal" stimuli. If you know
your Signal Detection Theory well, this may seem to you like a
perverse idea. If that is the case, please bear with me.

In this model each participant may have two different sets of
thresholds, one for "noise" and one for "signal" stimuli. Since the
thresholds are represented as random effects, we will account for the
fact that "random" thresholds may be correlated; They have to be
correlated, because if some threshold is higher (lower) than the
sample average, then the threshold above (below) it may also be higher
(lower).

@@

```{r echo=T,results="show"}                
fixed1 = list(eta = ~ 1, gamma = ~ -1 + stim)
random1 = list(list(group = ~ id, gamma = ~ -1 + stim))
## We have to make sure that stim is a factor for the separate
## intercepts and slopes parametrization to work as intended.
d2$stim = as.factor(d2$stim)
## Note that because for an ordinal model the stimulus variable (here
## the 'stim' variable) is not special, we have to include it in the
## additional variables vector
adata1 = aggregate_responses(d2, response = 'r', variables = c('id', 'stim'))
sdata1 = make_stan_data(adata1, fixed1, random1, model = 'ordinal', gamma_link = link)
model1 = make_stan_model(random1, model = 'ordinal', gamma_link = link)
if(fresh_start){
    fit1 = stan(model_code = model1, data = sdata1,
                pars = c('gamma_fixed', 'gamma_random_1', 'gamma_sd_1'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit1, '~/windows/temp/ordinal_fit_1.rds')
}else{
    fit1 = readRDS('~/windows/temp/ordinal_fit_1.rds')
}
## We will not look at the eta parameter since it was fixed at 0
print(fit1, probs = c(.025, .975), pars = c('gamma_fixed', 'gamma_sd_1'))
```

@@

Note the difference between the main criterion (row 4). This
difference approximates *d'*. Even though, as we will later see, this
model does not correctly account for the effect of the stimulus
(because it violates the non-negativity assumption), we can still
recover this effect pretty well, because *d'* is equivalent to a
constant shift of all the thresholds in one predefined
direction. Shifting all the criteria to the right is equivalent to
shifting the underlying distribution of "latent values" (or evidence)
to the left, so we have to subtract the average position of the
criteria for the "signal" stimuli from the average position of the
criteria for the "noise" stimuli. Here is our first attempt:

@@

In an SDT model we assume that the stimulus class affects *only* the
*d'* parameter, and that both criteria (gamma) and *d'* may differ
between the participants (or items, etc.). In bhsdtr *d'* is modelled
as *log(d') = delta*, which forces it to be non-negative. This is an
important assumption of Signal Detection Theory - if this assumption
is violated, for example, because of the response reversal in some
participants or conditions, then an SDT model is not valid and a
non-trivial generalization of SDT should be used instead (see the
bhsdtr preprint for a more detailed explanation and some
references). Otherwise, this is the same general hierarchical ordinal
regression model.

@@

```{r echo=T,results="show"}
adata2 = aggregate_responses(d2, stimulus = 'stim', response = 'r', variables = c('id'))
sdata2 = make_stan_data(adata2, fixed2, random2, model = 'sdt', gamma_link = link)
model2 = make_stan_model(random2, model = 'sdt', gamma_link = link)
if(fresh_start){
    fit2 = stan(model_code = model2, data = sdata2,
                pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1', 'gamma_sd_1',
                         'delta_random_1', 'gamma_random_1',
                         ## We will need this for the plot_sdt_fit function
                         'counts_new'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit2, '~/windows/temp/ordinal_fit_2.rds')
}else{
    fit2 = readRDS('~/windows/temp/ordinal_fit_2.rds')
}
print(fit2, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1', 'gamma_sd_1'))
```

So how well do the two *d'* estimates match? We can translate between
the delta and the *d'* parameters by using exponentiation. We have to
do this first, before doing anything else with the posterior samples,
because the exponential function is non-linear.

@@

The point and interval estimates match quite well, with some notable
exceptions. According to the ordinal model some credible intervals
around individual *d'* point estimates include 0 and one point
estimate is even below 0, which contradicts an important assumption of
the SDT model. We will address this issue later.

If the participant-specific *d'* estimates match, the sample average
*d'* should also match, right? Maybe the SDT model does not fit the
data well?

@@

```{r echo=T,results="show"}
delta2 = as.data.frame(t(apply(exp(s2[['delta_fixed[1,1]']] +
                                   s2[, grep('delta_random_1', names(s2))]), 2, sumfun)))
ggplot(data.frame(delta = delta2), aes(sample = delta)) +
    stat_qq() +
    ylab('Sample quantile of log(d\') = delta random effects') +
    xlab('Theoretical normal quantile') +
    labs(title = 'SDT model')
```

The individual *delta = log(d')* estimates in the SDT model seem to be
approximately normally distributed. We cannot plot *log(d')* estimates
based on the ordinal model, because according to this model one
participant has a negative *d'* value and even without this
participant's data there are quite a few negative posterior *d'*
samples. However, we can see if the *d'* point estimates based on the
ordinal model are approximately normally distributed:

@@

They seem to be approximately normally distributed as well. The
average / fixed effect *d'* in each model is estimated as the mean of
different distributions: in the ordinal model participant-specific
shifts are assumed to be normally distributed and in the SDT model
*delta = log(d')* random effects are assumed to be normally
distributed. We can easily account for this by making use of the
relationship between the parameters of the normal distribution and the
associated log-normal distribution:

@@

Before we go any further, just for fun, let's see what happens if we
introduce yet another assumption of selective influence, one that is
often accepted in Item Response Theory. We will fit a hierarchical SDT
model in which we will treat the thresholds as "item-parameters" and
we will treat the *d'* parameter as a "person-parameter".

```{r echo=T,results="show"}
fixed3 = list(delta = ~ 1, gamma = ~ 1)
random3 = list(list(group = ~ id, delta = ~ 1))
adata3 = aggregate_responses(d2, stimulus = 'stim', response = 'r', variables = c('id'))
sdata3 = make_stan_data(adata3, fixed3, random3, model = 'sdt', gamma_link = link)
model3 = make_stan_model(random3, model = 'sdt', gamma_link = link)
if(fresh_start){
    fit3 = stan(model_code = model3, data = sdata3,
                pars = c('delta_fixed', 'delta_sd_1', 'gamma_fixed', 'delta_random_1',
                         ## We will need this for the plot_sdt_fit function
                         'counts_new'),
                init_r = .5,
                iter = 5000,
                chains = 4)
    saveRDS(fit3, '~/windows/temp/ordinal_fit_3.rds')
}else{
    fit3 = readRDS('~/windows/temp/ordinal_fit_3.rds')
}
print(fit3, probs = c(.025, .975), pars = c('delta_fixed', 'gamma_fixed', 'delta_sd_1'))
```

@@

At the aggregated data level the fit is not perfect, but maybe it is
not that terrible either. What about the fit at the individual level?

@@

... we see that the simplified model not only gives biased
"person-parameter" estimates but the estimates seem more reliable
(here by about 20%). It does not make sense to look at the estimates
of the "item-parameters" since we already know that they have little
in common with the true thresholds.

Now back to Signal Detection Theory and the assumption of selective
influence. We can at least try to test this assumption. We will not
alter the ordinal model code to account for the non-negativity of *d'*
(and the log-normal distribution of random *d'* effects) just yet. As
we already know, even without this correction the estimates of
participants' shifts are similar to the *d'* estimates.

@@

Ok, we did simplify things, but it still makes sense to take a look at
the other gamma effects, because these other effects correspond to the
variability in the *pattern* of the thresholds, and the estimates of
the thresholds in our ordinal model seem perfectly fine.

@@

Note the clear pattern for the gamma_sd_1[.,2] parameters. Even though
the *sample-average* pattern of the thresholds seems to be different
for the two stimulus classes (gamma_fixed[.,2] except for
gamma_fixed[4,2] ~ d'), this difference in pattern seems to be fairly
constant across the participants: all the gamma_sd_1[.,2] parameters
seem to be near-zero, except for the gamma_sd_1[4,2] parameter, which
corresponds to the variability in *d'* values. This is weird, since
the pattern of the thresholds for the "noise" stimuli clearly varies
with participants - none of the 95% HPD intervals for the
gamma_sd_1[.,1] parameters include zero: in terms of standard
deviations of posterior distributions of standard deviations (phew)
most are in fact quite far from zero.

Now we will modify the stan model code to account for the
non-negativity of *d'*. Let's index the rows of our dataset by
*i*. Because of the way the model was parametrized...

@@

However, when stim = "signal" = 1, gamma[4]_i = gamma_fixed[4,1] +
gamma_random_1[g][4,1] + gamma_fixed[4,2] + gamma_random_1[g][4,2]. We
want this other part (the last sum which represents the difference in
gamma[4] between the "noise" and "signal" stimuli) to represent log of
average distance between the thresholds = log(shift) = log(d'), so we
have to subtract it from gamma[4]_i, exponentiate this sum, and add it
to gamma[4]_i.

@@

Interaction is tricky. For example, if you have an additive effect of
two factors on the original scale of the dependent variable...

@@

This means that whenever you see an interactive effect it may be an
artefact of the scale of the dependent variable. So, instead of
examining the effects in the gamma parameters. let's have a look at
the *average criteria / thresholds*. Notice that we have to add the
effect on gamma to gamma for stimulus = "noise" to obtain gamma for
stimulus = "signal".

@@

Now we can directly test if the average pattern of the criteria may be
different by comparing *centered* criteria vectors:

```{r echo=T,results="show"}
round(HPDinterval(as.mcmc((crit2 - apply(crit2, 1, mean)) - (crit1 - apply(crit1, 1, mean)))), 2)
```

When we center the criteria for the two classes of stimuli we no
longer see any evidence of the differences in their relative
positions. The only thing that seems to differ, and not by much, is
the main criterion, but this, I think, is confounded with *d'* in our
model. Selectivity seems to hold, kind of.

This is one way to *begin* to test an SDT model. An SDT model has
other problematic assumptions. One is that the time-accuracy tradeoff
can be ignored. If participants or conditions differ in the response
time, then the differences in *d'* may not mean that sensitivity is
different, because participants may simply take more time to respond
in certain experimental conditions, which inflates the *d'*
estimate. You need something like the diffusion model to deal with
this problem. If you combine the diffusion model with ratings, you
also need something like a hierarchical diffusion model with ratings
and an order-preserving link function.

The other problems that need to be addressed have to do with
non-linearity and aggregation. If you fit an SDT model, or any other
non-linear model, to data aggregated over grouping factors, such as
participants or items, because of non-linearity, which does not play
well with aggregation, some or all of the point and interval estimates
will be biased, perhaps severely so, and the inference will be
invalid. You can see how horribly biased the point *and* interval
estimates of every SDT parameter may become when you overly aggregate
your data in the bhsdtr preprint. By the way, I am now in the process
of examining the direction and magnitude of bias in SDT models fitted
to overly-aggregated data by using the excellent Confidence database
created by Doby Rahnev.

@@

Unfortunately, without additional assumptions, it is impossible to
model both kinds of variability at the same time even in the
hypothetical repeated-measurement-with-memory-wipeout-scenario,
because every shift in the internal value is indistinguishable from a
uniform effect in the thresholds, i.e., a constant shift of all the
thresholds. As I hope you already realize, the required additional
assumptions that make such models identifiable under certain
conditions are necessarily causal.

In many situations, this is both a statistical and a conceptual
problem. In particular, when the design is purely observational -
e.g., questionnaire measurement - without additional causal
assumptions, the way of describing the process of responding to a
Likert-type item in terms of internal values and thresholds is simply
redundant. As far as the design of the study is concerned, all we can
say about the internal value is that it is relative to the thresholds-
and the position of each threshold is relative to the positions of the
remaining thresholds and to the position of the internal
value. Without the selectivity assumption, to say that John thinks
that he was not crying at all during the last two weeks and to say
that *John's* threshold for the response "a few times" is above the
amount of crying that John thinks he experienced, or decided to
reveal, is to say the same thing.

It follows that *in such situations, we do not need the internal value
parameter at all*, unless there is some factor about which we can
perhaps assume that it can *only* produce a uniform effect in the
thresholds, such as the stimulus class ("noise" or "signal") in a
binary classification task, which can *perhaps* be assumed to induce a
change (*d'*) in the internal value only; In such cases the internal
value parameter may be justified and useful. We have obtained evidence
that this assumption perhaps holds for the analyzed dataset.

@@

It is usually not an *achievement* of Item Response Theory, it is just
an unrealistic *assumption*. The illusory nature of this achievement
becomes obvious as soon as we realize that without additional strong
causal assumptions it is impossible to distinguish between genuine
internal value effects and (dis)simulation (i.e., faking), which
conceptually corresponds to the change in the *participant*-specific,
not item-specific, decision thresholds, also known as
"item-parameters".

There is nothing in the *design* of the process of measurement by a
questionnaire that makes it possible to distinguish genuine
differences in internal values from the differences in the way of
responding. Because of random assignment of the stimulus class above
chance performance in a binary classification task cannot be faked,
and so the *d'* parameter definitely captures at least part of the
effect of the stimulus on the internal value. On the other hand, every
possible pattern of responses to a questionnaire can be faked.

The only way to (perhaps incorrectly) separate the "item" and "person"
effects is to *assume selective influence*, i.e., that there is some
factor f (such as the stimulus class in a binary classification task,
or, in case of a depression questionnaire, participants trait, or a
therapeutic intervention) which can only affect the internal value,
thus producing a constant shift in the thresholds. The assumption of
selective influence makes the model with the two kinds of parameters
identifiable.

@@

If we campared the estimates of the main decision criterion (i.e.,
bias) in SDT models fitted to two different large *samples* of
participants the difference would likely be negligible with very
narrow confidence / credible intervals, even though *individual
participants* clearly differ in how they use the main decision
criterion. By the same token, instability of item parameters in IRT
models cannot be estimated by assuming selective influence and
comparing estimates of "item-parameters" obtained in two different
large samples, because sample average estimates of item-parameters
*hide or misplace* (overdispersion) the intra- or inter-individual
variability in item parameters.

By assuming that internal values are participant-specific and item
parameters are item-specific we force the estimates of item parameters
to become estimates of average person-specific thresholds. This not
only begs the question, but also, because the model is non-linear,
makes the estimates of item-parameters *asymptotically biased*. It
does not take much thought to see that the assumption of constant
"item-parameters" at the group / sample / population level is utterly
unrealistic. It follows that the achievement of separating participant
and item effects in IRT models is often not an achievement at all, it
is just wishful thinking.

The primary justification of the selective influence assumption comes
from a psychological theory of the task, not from the design of the
study. All that is guaranteed by randomly assigning the stimulus class
(i.e., by choosing the stimulus class at random on every trial) is
that any statistical effect of the stimulus class is an unbiased
estimate of the total causal effect of the stimulus class. To
*empirically* justify the assumption of *selective* influence of the
stimulus class on the internal evidence value we have to *test* if the
stimulus class does not also affect the thresholds.
